% Determinacy Outline
\documentclass{article}


%preamble
%required
%\usepackage{Sweave} %Integrates R code with LaTeX for creating dynamic reports
\usepackage{natbib}%Provides citation and bibliography support
\usepackage{amsmath}%Enhances mathematical typesetting capabilities.
\usepackage{textcomp}%among other things, it allows degrees C to be added
\usepackage{float}%Helps with precise figure placement using the [H] option.
\usepackage[utf8]{inputenc} % allow funny letters in citations 
\usepackage[nottoc]{tocbibind} %should add Re fences to the table of contents?
\usepackage{amsmath} % making nice equations 
\usepackage{listings} % add in stan code
\usepackage{xcolor}
\usepackage{capt-of}%allows me to set a caption for code in appendix 
\usepackage[export]{adjustbox} % adding a box around a map
\usepackage{lineno}
\linenumbers

% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as odd side margin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
%\fancyhead[LO]{Frederik Baumgarten}
%\fancyhead[RO]{Research Proposal}
% more optionals! %

\usepackage{graphicx}
\graphicspath{{/Users/frederik/github/PlantDeterminism/figures/}} % specify the path to your figures directory




\begin{document}
	\date{\today}
	
	\title{Invest now, get paid later? Growth strategies to cope with environmental stress and benefit from extended growing seasons in a future climate}
\maketitle

\author{Frederik Baumgarten\textsuperscript{1,2}, potentially with Yann Vitasse, EM Wolkovich, Sally and Rob)}


%What is determinis and why does it matter (tree growth  x CC GSL and environmental stress)

\section{1. Intro}
\subsection{Investment by trees}
Why it matters with cc opening paragraph
\subsection{Introduce seasonality}
frost etc. investment in a risky environment
\subsection{introduce soil moisture}
can we forecast with CC from temp and moisture alone
env. drivers of groth, but this is not enough
instrinic programming matters

\section{2. Introduce Concept of deerminism}
history , old poplar in 70ies, many terms
what is it? fig
1 sentense represented as dichotomous but intermediate examples exist

\section{3. what controls/drivers of determinism}
a)total physiology/envrironmental control 1-2 para
	warm, wet, ideal leads to indeterminate?
	but environment varies above/belowground
	R:S matters
	but this perspectives suggest you can flip species strategy completely and we do not see that.
b) Genetically controlled /species variation see box/table/fig with list of species
general overview of species vary
c) these are fundamental trade-offs - both successful and co -occur in communities
successional stage, ontogeny, life span

...but will both still be successful with CC?

\section{4. CC }
Figure explain. temp vs. moisture vs. who wins/ phenology
C sequestration
\section{5. future directions}
Questions critical forcasting now to larger evolution questions

a) (1-2 paragraphfs) how much does dezerminism predict bufferin vs. exploitation of env. change?
explain this in simple words
experiment phaenoflex

b) How flexible are spp. in determinism?
need an answer to get anywhere but must approach from several fields

bi) Physiology: universal across meristmes? linked to xylogenesis?
universal across allocation? (reproduction, storage and growth?)
gene/hormones (1-2 para)
bii) Evol. history (end on metric fortshadowing)

c) better metricx
opening....its treated dichotomy by 1 field, flexible by a different so clearly it has variability we need to understand to move forward

we propose several metrics from best to worse, dificult to easy.
(1) nleaves  EOS /n leaves in buds SOS

(2) shoot elongation/dendrometers radial growth? 

(3) "second flushes" in large databases? US national forest network 
NDVI second flushes?
 
\end{document}


